### 2014-04-29 doc/coppermtn_multilevel/tech_report/multilevel_mc.tex
finished first draft of copper multilevel slides
adding some heterogenous poisson results to the multilevel mc note

Recent work in Monte Carlo methods for problems in finance and
stochastic partial differential equations has indicated that the
computationally complexity of the problem can be dramatically reduced
by incorporating multigrid concepts into the solution scheme
\cite{heinrich_2001,giles_2008,cliffe_2011}. In this work, we adapt
those ideas and apply them to the Monte Carlo problem for linear
systems as a means of reducing the computational complexity of the
algorithm. To begin, we first introduce the elliptic model problem for
our numerical experiments and compare the spectral behavior of the
Monte Carlo method to traditional iterative smoothers that would be
used with the multigrid method. Next, we present the multilevel Monte
Carlo method for linear systems using a general algebraic
formulation. Finally, we present results using the model problem the
demonstrate the superiority of the multilevel method in terms of time
to solution using both geometric and algebraic multigrid
representations of the linear problem.

@@

    \For{ l = L...1 }
    \State $\ve{d}_{l-1} = ( \ve{I} + \ve{P}_{l} ) \ve{d}_{l}$
    \Comment{Collapse the tallies to the finest grid}
    \EndFor

@@

%%---------------------------------------------------------------------------%%
\section{Algebraic Multigrid Results}
\label{sec:algebraic_results}

>>>>

Recent work in Monte Carlo methods for problems in finance and
stochastic partial differential equations has indicated that the
computationally complexity of the problem can be dramatically reduced
by incorporating multigrid concepts into the solution scheme
\cite{heinrich_2001,giles_2008,cliffe_2011}. In this work, we adapt
those ideas and apply them to the Monte Carlo problem for linear
systems as a means of reducing the computational complexity of the
algorithm. To begin, we first introduce the elliptic model problem for
our numerical experiments and compare the spectral behavior of the
Monte Carlo method to traditional iterative smoothers that would be
used with the multigrid method. Finally, we present results using the
model problem the demonstrate the superiority of the multilevel method
in terms of time to solution using a geometric multigrid
representations of the linear problem.

@@

    \For{ l = L...1 }
    \State $\ve{d}_{l-1} = \ve{d}_{l-1} + \ve{P}_{l} \ve{d}_{l}$
    \Comment{Collapse the tallies to the finest grid}
    \EndFor

@@

\subsection{Heterogeneous Results}
We next solve the same linear problem as before using $M=2$ but this
time with a constant forcing of $\ve{b}=1.0$ such that:
\begin{equation}
  \ve{A}\ve{x} = \ve{b}\:.
  \label{eq:heterogeneous_problem}
\end{equation}
We will choose an initial guess of $\ve{0}$ for these calculations as
the solution is non-trivial. Studies indicate that residual norms are
controlled by the convergence of the finest level. However, adding
additional levels provides significant smoothing to the statistical
error. For example, given a fixed amount of compute time of
approximately 7 seconds, Eq~(\ref{eq:heterogeneous_problem}) was
solved using 1 level and 10,000 samples and again using 9 levels and
1,546,783 total samples ($N=1,000,000$). The 1 level calculation gave
$||\ve{r}||_\infty = 0.027$ and $||\ve{r}||_2 = 0.230$ while the 1
level calculation gave $||\ve{r}||_\infty = 0.129$ and $||\ve{r}||_2 =
0.812$. Based on the residual norm values, it would seem as though the
1 level result was superior as better convergence was achieved in the
same amount of compute time. However, the structure of the actual
solutions would indicate otherwise. Figure~\ref{fig:7_sec_1_level}
gives the 1 level result and Figure~\ref{fig:7_sec_9_level} gives the
9 level result, both plotted against an exact solution computed using
a minimal residual iteration. Clearly, the estimate produced by the 9
level computation is more representative of the exact solution than
the single level scheme due to the well-converged results provided
coarse level estimates.
\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=4in]{7_sec_1_level.pdf}
  \end{center}
  \caption{\textbf{Solution of the Poisson problem using the geometric
      multilevel Monte Carlo method with 1 level and 6.87 seconds of
      compute time.}}
  \label{fig:7_sec_1_level}
\end{figure}
\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=4in]{7_sec_9_level.pdf}
  \end{center}
  \caption{\textbf{Solution of the Poisson problem using the geometric
      multilevel Monte Carlo method with 9 levels and 6.78 seconds of
      compute time.}}
  \label{fig:7_sec_9_level}
\end{figure}
To show that the 1 level calculation will eventually converge to the
correct result, Figure~\ref{fig:converged_1_level} shows the solution
with 1,000,000 samples. This calculation required 665.7 seconds to run
and gave $||\ve{r}||_\infty = 0.0025$ and $||\ve{r}||_2 = 0.022$.
\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=4in]{converged_1_level.pdf}
  \end{center}
  \caption{\textbf{Solution of the Poisson problem using the geometric
      multilevel Monte Carlo method with 1 level and 665.7 seconds of
      compute time.}}
  \label{fig:converged_1_level}
\end{figure}

Clearly, these results indicate that quality of solution from the
multilevel method may not lie entirely in a residual norm metric as
they are global in nature. For the cases above we see that the
multilevel estimator has signficantly better local convergence
properties for a fixed amount of work when compared to single level
methods. Therefore, we must determine if this improved local
convergence is satisfactory to improve the convergence of MCSA or if
convergence of global metrics is also required of the Monte Carlo method.
